---
title: "Exercise 2: From models to forecasts"
author: "Alexey Shiklomanov"
---

**Problem 1: Generate plots of the logistic growth model at r = 1.95, 2.05, 
2.5, and 2.8 Describe the trajectory observed in each case.**

Define logistic function.
```{r}
logistic <- function(r, K, n0, NT){
    n <- rep(n0, NT)
    for(t in 2:NT){
        n[t] <- n[t-1] + r*n[t-1] * (1-n[t-1]/K)
    }
    return(n)
}
```

Set parameters
```{r}
rvec <- c(1.95, 2.05, 2.5, 2.8)
K <- 10
n0 <- 0.1
NT <- 30
```

Create plots.
```{r}
par(mfrow=c(2,2))
for(r in rvec){
    tt <- 1:NT
    n <- logistic(r, K, n0, NT)
    main <- sprintf("r = %.2f", r)
    plot(tt, n, type='l', main=main,
         xlab="Time", ylab="Population")
}

```

In each case, the plot grows rapidly early on before reaching values around 
*K*.  At that point, the plot oscillates depending on the value of *r*. For low 
values, the oscillations are relatively muted, and for the lowest *r* (1.95), 
population converges to a stable state. For the second lowest *r*, the 
population oscillates evenly and apparently forever around *K*. At increasingly 
large values of *r*, the oscillations grow in magnitude and become less even.


**Problem 2: Choose another probability distribution and generate graphs of 
the probability density function, the cumulative distribution function, the 
quantile function, and a histogram of samples from that distribution.**


For this exercise, I will use the gamma. Because I'm all about that Gibbs 
sampling for normal distrubions with known means.

```{r}
# Set gamma parameters
shape <- 3
scale.p <- 5
xx <- seq(0, 60, by=1)
qq <- seq(0, 1, by=0.05)
nsamp <- 1e5
par(mfrow=c(2,2), mar=c(2,6,2,2))
plot(xx, dgamma(xx, shape, scale=scale.p), type='l',
     xlab="X", ylab="P(X)", main="Gamma PDF")
plot(xx, pgamma(xx, shape, scale=scale.p), type='l',
     xlab="X", ylab=expression(integral(P(X)*dx, -infinity, X)), main="Gamma CDF")
plot(qq, qgamma(qq, shape, scale=scale.p), type='l',
     xlab="X", ylab=expression(CDF^-1), main="Gamma Q")
hist(rgamma(nsamp, shape, scale=scale.p), xlab="X", ylab="Counts",
     main=sprintf("Gamma samples (n = %d)", nsamp))
```

**Problem 3 Numerically transform a lognormal(meanlog=0,sdlog=0.5) through 
sin(x) using Monte Carlo simulation. Include histograms of the original and 
transformed distributions. Report the mean, median, and 95% CI for both 
distributions and indicate these values on the histograms.**

Define stats and plotting function
```{r}
statplot <- function(samples, ...){
    s.mean <- mean(samples)
    s.q <- quantile(samples, c(0.025, 0.5, 0.975))
    hist(samples, ...)
    abline(v = s.q, lty=c(2,1,2), lwd=3, col="orange")
    abline(v = s.mean, lty=3, lwd=3)
}
```

```{r}
mu <- 0
sigma <- 0.5
nsamp <- 1e5
lnorm <- rlnorm(nsamp, mu, sigma)
lnorm.sin <- sin(lnorm)
par(mfrow=c(1,2))
statplot(lnorm, main="Lognormal samples")
statplot(lnorm.sin, main="Sin(lognormal) samples")
```

# Parameter error

First, set the parameter variabilities.
```{r}
r <- 1.95
K <- 10
n0 <- 0.1
NT <- 30
r.sd = 0.2     ## standard deviation on r
K.sd = 1.0     ## standard deviation on K
NE = 1000      ## Ensemble size
```

Then, run the ensemble analysis. Note that for high *r* values, the model 
sometimes generates negative values, which are impossible for populations.  
Therefore, I've constained the model to have a hard minimum at 0, which 
indicates the population has undergone total collapse.

```{r}
n = matrix(n0,NE,NT)   # storage for all simulations
rE = rnorm(NE,r,r.sd)  # sample of r
KE = rnorm(NE,K,K.sd)  # sample of K
for(i in 1:NE){        # loop over samples
    n[i,] <- logistic(rE[i], KE[i], n0, NT)
}
n[n < 0] <- 0
n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
```

**Problem 4: Plot histograms of the samples of r and K used for the 
simulation.**

```{r}
par(mfrow=c(1,2))
hist(rE, main="Samples of r")
hist(KE, main="Samples of KE")
```

**Problem 5: Plot a sample of 10 different trajectories from your ensemble (on 
one graph).**

```{r}
n.rows <- sample(1:1000, 10)
n.plot <- n[n.rows,]
matplot(t(n.plot), type='l', xlab='Time', ylab="Populaiton",
        main="10 realizations of logistic model")
```

**Problem 6: Plot a histogram of your population forecast at time = 15.**

```{r}
hist(n[,15], main="Population at t=15")
```

**Problem 7: Plot the median trajectory and 95% CI.**

Base R graphics doesn't have an intuitive way of doing transparency, so I have 
to plot the `ciEnvelope` first. But, I can't call `polygon` on a plot that 
doesn't exist yet. So, I first set up a blank plot (`type='n'`) with the right 
specifications, then plot the confidence envelope on it, and finally draw the 
median line on top of everything.

```{r}
ciEnvelope <- function(x,ylo,yhi,col="lightgrey",...){
    xcoords <- c(x, rev(x), x[1])
    ycoords <- c(ylo, rev(yhi), ylo[1])
  polygon(xcoords, ycoords, border = NA, col=col,...) 
}
plot(0, type='n', xlab='Time', ylab='Population',
     main = 'Median and 95% CI of logistic model',
     xlim=c(0,NT), ylim=range(n.stats))
ciEnvelope(1:NT, n.stats[1,], n.stats[3,])
lines(n.stats[2,], type='l')
```

**Problem 8: Add a 50% CI (i.e. 25% to 75%) to the plot. Note that you'll have 
to both compute the summary statistics for this interval and plot the envelope 
in a different color.**

```{r}
n.stats2 <- apply(n, 2, quantile, c(0.025, 0.25, 0.5, 0.75, 0.975))
plot(0, type='n', xlab='Time', ylab='Population',
     main = 'Median and 50% and 95% CIs of logistic model',
     xlim=c(0,NT), ylim=range(n.stats))
ciEnvelope(1:NT, n.stats2[1,], n.stats2[5,], col="lightgrey")
ciEnvelope(1:NT, n.stats2[2,], n.stats2[4,], col="red1")
lines(n.stats2[3,], type='l', lwd=2)
```

# Extra credit: Initial conditions

```{r}
ic.analysis <- function(r){
    n0.sd = 0.6
    n0s = rlnorm(NE,log(n0),n0.sd)
    n = matrix(n0s,NE,NT)
    for(i in 1:NE){
        n[i,] <- logistic(r, K, n0s[i], NT)
    }
    n[n < 0] <- 0
    n.stats <- apply(n, 2, quantile, c(0.025, 0.5, 0.975))
    plot(0, type='n', xlab='Time', ylab='Population',
        main = sprintf('r = %.2f', r),
        xlim=c(0,NT), ylim=range(n.stats))
    ciEnvelope(1:NT, n.stats[1,], n.stats[3,])
    lines(n.stats[2,], type='l')
}
```

**Problem 9: Plot the median & 95% interval.**

```{r}
r = 0.3
ic.analysis(r)
```

**Problem 10: Repeat with r equal to 1.95, 2.05, and 2.8.**

```{r}
par(mfrow=c(2,2))
for(r in rvec){
    ic.analysis(r)
}
```
